---
title: "Harris, Acquisition of surgical skills"
author: "Peter Geelan-Small"
date: "26th July, 2021"
output: html_notebook
---


```{r message = F, warning = F}

library(tidyverse)
library(RcmdrMisc)
library(emmeans)
library(effects)
library(statmod)
library(sjPlot)

```



```{r}

surg <- read.csv("harris acquisition surgical skills 2017.csv")

str(surg)

```

Recode "Condition" with names of groups.

```{r}

surg <-
    surg %>%
    mutate(Condition2 = ifelse(Condition == 1, "Novice", 
        ifelse(Condition == 2, "Expert", 
            ifelse(Condition == 3, "Mixed", "Control"
            )
        )
    )
    )
    
surg

```




# Mean jerk by trial and group

Make appropriate data set

First analysis had Condition consisting o baseline, post-intervention and retention. It might be better to use baseline as a predictor, as there was no training done at baseline. The second lot of analyses are done in this way. No data values have been excluded.



```{r}

jerk_mean1 <- 
    surg %>%
    select(Participant, MeanJerk1, MeanJerk2, MeanJerk3) %>%
    pivot_longer(!Participant, names_to = "Trial", 
    values_to = "MeanJerk")

#head(jerk_mean1) 

jerk_mean1 <- 
    jerk_mean1 %>%
    mutate(Trial2 = ifelse(Trial == "MeanJerk1", "Baseline",
                           ifelse(Trial == "MeanJerk2", "Post", 
                                  "Retention")))

#head(jerk_mean1) 

## Add "Condition2"

condit3 <- rep(surg$Condition2, each = 3)

jerk_mean2 <- data.frame(jerk_mean1, Condition2 = condit3)

#jerk_mean2[1:30, ]

jerk_mean2$Participant <- factor(jerk_mean2$Participant)

jerk_mean2$Trial2 <- factor(jerk_mean2$Trial2)

jerk_mean2$Condition2 <- factor(
    jerk_mean2$Condition2, 
    levels = c("Novice", "Expert", "Mixed", "Control"))

rm(jerk_mean1)
    
```



```{r}

summary(jerk_mean2)

```


```{r}

ggplot(jerk_mean2, aes(x = "", y = MeanJerk)) +
  geom_boxplot(outlier.shape = NA, width = 0.4) +  # avoid plotting outliers twice
  geom_jitter(position = position_jitter(width = 0.1, height = 0))

```


```{r}

ggplot(jerk_mean2, aes(x = Trial2, y = MeanJerk)) +
  geom_boxplot(outlier.shape = NA, width = 0.4) +  # avoid plotting outliers twice
  geom_jitter(position = position_jitter(width = 0.1, height = 0))
    
```



```{r}

ggplot(jerk_mean2, aes(x = Condition2, y = MeanJerk)) +
  geom_boxplot(outlier.shape = NA, width = 0.4) +  # avoid plotting outliers twice
  geom_jitter(position = position_jitter(width = 0.1, height = 0))
 
```



```{r}

jerk_mean2 <- 
    jerk_mean2 %>%
    mutate(logMeanJerk = log(MeanJerk))

```



```{r eval = F}

tapply(jerk_mean2$MeanJerk, INDEX = jerk_mean2$Trial2, sd, na.rm = T)

```



```{r eval = F}

jerk_mean2$TrialCond <- interaction(jerk_mean2$Trial2, 
                                    jerk_mean2$Condition2, 
                                    sep = "_")

ggplot(jerk_mean2, aes(x = TrialCond, y = logMeanJerk, colour = TrialCond)) +
  geom_boxplot(outlier.shape = NA) +  # avoid plotting outliers twice
  geom_point(alpha = 0.5, size = 1, position = position_jitterdodge()) +
  labs(x = "Time-Condition", y = "Jerk (m/s^3)", colour = "TrialCond") +
  #guides(fill = guide_legend(title = "Condition")) +
  ggtitle("Mean jerk - log-transformed data") +
  theme_classic() 

```



```{r}

ggplot(jerk_mean2, aes(x = "", y = logMeanJerk)) +
  geom_boxplot(outlier.shape = NA, width = 0.4) +  # avoid plotting outliers twice
  geom_jitter(position = position_jitter(width = 0.1, height = 0))


```






```{r}

jerk.lme1 <- lme(MeanJerk ~ Condition2 * Trial2, 
                 random = ~ 1 | Participant,
                 data = jerk_mean2,
                 na.action = "na.omit")

anova(jerk.lme1)

```



```{r}

jerk.lme1a <- lme(MeanJerk ~ Trial2 * Condition2, 
                 random = ~ 1 | Participant,
                 data = jerk_mean2,
                 na.action = "na.omit")

anova(jerk.lme1a)

```



```{r}

plot(jerk.lme1)  
## Default is innermost level of grouping, 
##   viz. not population level, but level "1"

```



```{r}

plot(jerk.lme1, 
     sqrt(abs(resid(., type = "p"))) ~ fitted(.), 
     type = c("p", "smooth"))

```



```{r}

## Need to use Pearson residuals calculated conditional on random 
##   effects - see Pinheiro & Bates (i.e. level = 1 which is default)

jerk.sres10 <-  residuals(jerk.lme1, level = 0, type = "pearson")

jerk.sres11 <-  residuals(jerk.lme1, level = 1, type = "pearson")

qqnorm(jerk.sres10)
qqline(jerk.sres10)

```

The preferred residual type is plotted below.

```{r}

qqnorm(jerk.sres11)
qqline(jerk.sres11)

```




```{r}

jerk.lme2 <- lme(MeanJerk ~ Condition2 + Trial2, 
                 random = ~ 1 | Participant,
                 data = jerk_mean2,
                 na.action = "na.omit")

anova(jerk.lme2)

```


```{r}

jerk_mean_emm1 <- emmeans(jerk.lme1, ~ Condition2 | Trial2)

summary(jerk_mean_emm1)

```


```{r}

jerk_mean_emm2 <- emmeans(jerk.lme2, pairwise ~ Trial2)

summary(jerk_mean_emm2)

```


```{r}

jerk_mean_emm1a <- emmeans(jerk.lme1, ~ Trial2 | Condition2)

plot(jerk_mean_emm1a, horizontal = F)

```



```{r}

plot(allEffects(jerk.lme1), lines = list(multiline = T), 
     confint = list(style = "bars"))
     
```



```{r}

plot(predictorEffects(jerk.lme1, ~ Trial2), lines = list(multiline = T), 
     confint = list(style = "bars"))

```



```{r}

jerk.lme3 <- lme(logMeanJerk ~ Condition2 * Trial2, 
                 random = ~ 1 | Participant,
                 data = jerk_mean2,
                 na.action = "na.omit")

anova(jerk.lme3)

```



```{r}

plot(jerk.lme3)

```



```{r}

AIC(jerk.lme1); AIC(jerk.lme2) 

```


<br>

## KT time & score


```{r}

surg$Gender <- factor(surg$Gender)

surg$Condition2 <- factor(surg$Condition2)

surg$Handedness <- factor(surg$Handedness)

```


```{r}

ggplot(surg, aes(x = "", y = KTTime)) +
  geom_boxplot(outlier.shape = NA, width = 0.4) +  # avoid plotting outliers twice
  geom_jitter(position = position_jitter(width = 0.1, height = 0))


```


```{r}

ggplot(surg, aes(x = Gender, y = KTTime)) +
  geom_boxplot(outlier.shape = NA, width = 0.4) +  # avoid plotting outliers twice
  geom_jitter(position = position_jitter(width = 0.1, height = 0))

```



```{r}

ktt.lm1 <- lm(KTTime ~ Condition2 * Gender, 
              data = surg)

anova(ktt.lm1)

```


```{r}

ktt.lm2 <- lm(KTTime ~ Condition2 + Gender, 
              data = surg)

anova(ktt.lm2)

```


```{r}

summary(ktt.lm2)

```


```{r}

ktt.emm2 <- emmeans(ktt.lm2, ~ Gender)

summary(ktt.emm2)

```


```{r}

plot(ktt.lm2, which = 1)

```



```{r}

plot(ktt.lm2, which = 2)

```



```{r}

surg$logKTTime <- log(surg$KTTime)

ggplot(surg, aes(x = "", y = logKTTime)) +
  geom_boxplot(outlier.shape = NA, width = 0.4) +  # avoid plotting outliers twice
  geom_jitter(position = position_jitter(width = 0.1, height = 0))

```



```{r}

ggplot(surg, aes(x = Gender, y = logKTTime)) +
  geom_boxplot(outlier.shape = NA, width = 0.4) +  # avoid plotting outliers twice
  geom_jitter(position = position_jitter(width = 0.1, height = 0))

```



```{r}

ktt.lm3 <- lm(logKTTime ~ Condition2 + Gender , 
              data = surg)

anova(ktt.lm3)

```


<br>

```{r}

plot(ktt.lm3, which = 1)

```



```{r}

plot(ktt.lm3, which = 2)

```

## KT score

```{r}

ggplot(surg, aes(x = "", y = KTscore)) +
  geom_boxplot(outlier.shape = NA, width = 0.4) +  # avoid plotting outliers twice
  geom_jitter(position = position_jitter(width = 0.1, height = 0.1))


```



```{r}

ggplot(surg, aes(x = Condition2, y = KTscore)) +
  geom_boxplot(outlier.shape = NA, width = 0.4) +  # avoid plotting outliers twice
  geom_jitter(position = position_jitter(width = 0.1, height = 0.1))

```



```{r}

kts.lm1 <- lm(KTscore ~ Condition2 * Gender, 
              data = surg)

anova(kts.lm1)

```


```{r}

plot(kts.lm1, which = 1)

```



```{r}

plot(kts.lm1, which = 2)

```


```{r}

kts.lm2 <- lm(KTscore ~ Condition2 + Gender , 
              data = surg)

anova(kts.lm2)

```


```{r}

plot(kts.lm2, which = 1)

```



```{r}

plot(kts.lm2, which = 2)

```




```{r}

summary(kts.lm2)

```


```{r}

kts.emm2 <- emmeans(kts.lm2, pairwise ~ Gender)

summary(kts.emm2)

```


```{r}

confint(kts.emm2)$contrasts

```


```{r}

contrast(kts.emm2, "eff")

```



```{r}

pairs(kts.emm2)

```



```{r}

kts.lm3 <- lm(KTscore ~ Condition2, data = surg)

anova(kts.lm3)

```


```{r}

res_fit_kts3 <- data.frame(res_kts3 = rstandard(kts.lm3),
                           fit_kts3 = fitted(kts.lm3))

ggplot(res_fit_kts3, aes(x = fit_kts3, y = res_kts3)) +
  geom_jitter(width = 0.01, height = 0.01)

#plot(kts.lm3, which = 1)

```


```{r}

plot(kts.lm3, which = 2)

```



In original scoring scheme, max. score is 13 and acceptable score is 10. Make binary variable here pass/fail corresponding to this.



```{r}

numSummary(surg$KTscore)

```


```{r}

surg$KTscore13 <- floor(surg$KTscore / 8 * 13)

cbind(surg$KTscore, surg$KTscore13)

surg <-
  surg %>%
  mutate(KTskill = ifelse(KTscore13 < 10, "Unsat", "Sat"))

#cbind(surg$KTscore13, surg$KTskill)

surg$KTskill <- factor(surg$KTskill)

```



```{r}

KTscore_tbl <- with(surg, table(Condition2, KTskill))

KTscore_tbl

```


```{r}

KTscore_tbl_st <-
  surg %>%
  select(Condition2, KTskill) %>%
  group_by(Condition2, KTskill) %>%
  na.omit() %>%
  summarise(n_obs = n())

KTscore_tbl_st2 <- as.data.frame(KTscore_tbl)
                                 
surg %>%
  filter(!is.na(KTskill) == T) %>% 
  ggplot(., aes(Condition2)) +
  geom_bar(aes(fill = KTskill))
  
```


Above plot possibly not as useful as one below.


```{r}

cond_skill <- surg %>%
  group_by(Condition2, KTskill) %>%
  summarise(n_obs = n()) %>%
  na.omit() 

N <- sum(cond_skill$n_obs)

ggplot(cond_skill, aes(x = Condition2, y = n_obs, fill = KTskill)) + 
  geom_bar(position = "fill", stat = "identity") +
#  scale_fill_manual(values = col.pal) +
  labs(y = "Proportion")


```


```{r}

chisq.test(KTscore_tbl, correct = F)

```


<br>

## Errors per sec. and mean jerk at baseline


```{r}

ggplot(surg, aes(x = MeanJerk1, y = ERRORS.SEC_RT1)) +
    geom_point()

```



```{r}

ggplot(surg, aes(x = "", y = ERRORS.SEC_RT1)) +
  geom_boxplot(outlier.shape = NA, width = 0.4) +  # avoid plotting outliers twice
  geom_jitter(position = position_jitter(width = 0.1, height = 0.1))

```



```{r}

ggplot(surg, aes(x = "", y = MeanJerk1)) +
  geom_boxplot(outlier.shape = NA, width = 0.4) +  # avoid plotting outliers twice
  geom_jitter(position = position_jitter(width = 0.1, height = 0.1))

```


Make data subset removing NAs from ERRORS.SEC_RT1, MeanJerk1, RT1Errors and RT1Time.

```{r}

surg_sub1 <-
  surg %>%
  select(ERRORS.SEC_RT1, MeanJerk1, RT1Errors, RT1Time) %>%
  na.omit()

numSummary(surg_sub1)

```



```{r}

err.lm1 <- lm(ERRORS.SEC_RT1 ~ MeanJerk1, data = surg_sub1)

anova(err.lm1)

```



```{r}

summary(err.lm1)

```



```{r}

plot(err.lm1, which = 1)

```


Not satisfactory!


```{r}

plot(rstandard(err.lm1) ~ fitted(err.lm1))
abline(h = 0, lty = 3)
     
```



```{r}

plot(err.lm1, which = 2)

```


```{r}

plot(err.lm1, which = 4)

```

Remove data points with std. resids > 3.


```{r}

surg_sub1$sres_err1 <- rstandard(err.lm1)

surg_sub2 <-
  surg_sub1 %>%
  filter(sres_err1 < 3) 
  
dim(surg_sub2)

```



```{r}

err.lm1a <- lm(ERRORS.SEC_RT1 ~ MeanJerk1, data = surg_sub2)

summary(err.lm1a)

```


```{r}

plot(rstandard(err.lm1a) ~ fitted(err.lm1a))
abline(h = 0, lty = 3, col = "red")

```



```{r}

plot(err.lm1a, which = 2)

```



```{r}

plot(err.lm1a, which = 4)

```

<br>

Try a log transformation of error rate

```{r}

surg_sub1$logERRORS.SEC_RT1 <- log(surg_sub1$ERRORS.SEC_RT1 )

ggplot(surg_sub1, aes(x = "", y = logERRORS.SEC_RT1)) +
  geom_boxplot(outlier.shape = NA, width = 0.4) +  # avoid plotting outliers twice
  geom_jitter(position = position_jitter(width = 0.1, height = 0.1))


```




```{r}

err.lm2 <- lm(logERRORS.SEC_RT1 ~ MeanJerk1, data = surg_sub1)

anova(err.lm2)

```



```{r}

plot(err.lm2, which = 1)

```



```{r}

plot(err.lm2, which = 2)

```



```{r}

plot(err.lm1, which = 3)

```


```{r}

newX <- seq(from = min(surg_sub1$MeanJerk1, na.rm = T),
             to = max(surg_sub1$MeanJerk1, na.rm = T),
             length = 100)

newMJ <- data.frame(MeanJerk1 = newX)

```




```{r}

err_RT1_pred1 <- predict(err.lm1, newdata = newMJ, interval = "confidence")

yhat1 <- data.frame(err_RT1_pred1, MeanJerk = newMJ$MeanJerk1)

```




```{r}

plot(fit ~ MeanJerk, data = yhat1, type = "l", ylim = c(0, 0.4),
     ylab = "ERRORS.SEC_RT1")
points(surg_sub1$MeanJerk1, surg_sub1$ERRORS.SEC_RT1)
lines(yhat1$MeanJerk, yhat1$lwr, lty = 2)
lines(yhat1$MeanJerk, yhat1$upr, lty = 2)

```


<br>


```{r}

summary(err.lm2)

```



```{r}

confint(err.lm2)

```



```{r}

err_RT1_pred2 <- predict(err.lm2, newdata = newMJ, interval = "confidence")

err_RT1_pred_bt2 <- exp(err_RT1_pred2)

yhat2 <- data.frame(err_RT1_pred2, MeanJerk = newMJ$MeanJerk1)

yhat_bt2 <- data.frame(err_RT1_pred_bt2, MeanJerk = newMJ$MeanJerk1)

```



```{r}

plot(fit ~ MeanJerk, data = yhat2, type = "l", ylim = c(-5, 0),
     ylab = "logERRORS.SEC_RT1")
points(surg_sub1$MeanJerk1, surg_sub1$logERRORS.SEC_RT1)

```



```{r}

plot(fit ~ MeanJerk, data = yhat_bt2, type = "l", ylim = c(0, 0.4))
points(surg_sub1$MeanJerk1, surg_sub1$ERRORS.SEC_RT1)

```




<br>


Use Poisson GLM with number of errors as response and time to completion as offset.


```{r}

ggplot(surg_sub1, aes(x = MeanJerk1, y = RT1Errors)) +
    geom_point()

```



```{r}

err.glm1 <- glm(RT1Errors ~ MeanJerk1 + offset(log(RT1Time)),
                family = poisson, data = surg_sub1)

summary(err.glm1)

```



```{r}

err.qres.p <- qresiduals(err.glm1)

qqnorm(err.qres.p)
abline(0, 1)

```



```{r}

plot(qresiduals(err.glm1) ~ fitted(err.glm1))
abline(h = 0)

```



```{r}

#ref_grid(err.glm1)

err.emm1 <- emmeans(err.glm1, "MeanJerk1", type = "response", offset = 0)

summary(err.emm1)

```


<br>


```{r}

err.glm2 <- glm.nb(RT1Errors ~ MeanJerk1 + offset(log(RT1Time)),
                   data = surg_sub1)

summary(err.glm2)

```



```{r}

err.qres.nb <- qresiduals(err.glm2)

qqnorm(err.qres.nb)
abline(0, 1)

```



```{r}

plot(qresiduals(err.glm2) ~ fitted(err.glm2))
abline(h = 0)

```





```{r}

err.glm3 <- glm(RT1Errors ~ MeanJerk1 + offset(log(RT1Time)), 
                family = quasipoisson,
                data = surg_sub1)

summary(err.glm3)

```



```{r}

plot(err.glm3, which = 1)

```



```{r}

plot(err.glm3, which = 2)

```



```{r}

plot(err.glm3, which = 4)

```



```{r}

AIC(err.glm1)

AIC(err.glm2)

```





Save all data objects for convenience

```{r}

save(KTscore_tbl, KTscore_tbl_st, surg, surg_sub1, surg_sub2, 
     file = "harris_data_obj.RData")

```



```{r}



```

